---
title: "Report CORIANDR: ChrOmosomal abeRration Identifier AND Reporter in R"
date: "`r format(Sys.time(), '%d-%m-%Y')`"
output: pdf_document
params:
  data: patient
---


```{r setup, include=FALSE}
#knitr::opts_knit$set(root.dir = "/home/vera/dissertation/test.pat.munich/")
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
# knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
knitr::opts_chunk$set(dev = 'pdf')

library(knitr)
```


```{r patient meta data, echo = FALSE}
patient.meta = data.frame(t(read.table(paste0(params$data, ".meta.tsv"), stringsAsFactors = F, row.names=1)))
if(tolower(patient.meta$gender) %in% c("f", "female")) {
  patient.meta$gender = "f"
} else {
  patient.meta$gender = "m"
}
```

## Whole genome sequencing depth of patient data

```{r data definition, echo = FALSE, fig.width=7, fig.height=3}
# Read patient counts from featureCounts output
patient = read.table(as.character(patient.meta$count_data), header=TRUE, row.names = 1, stringsAsFactors = F)
patient = patient[patient$Chr != 'chrM', ]

# Read PON table
pon = read.table(as.character(patient.meta$pon), header=TRUE, row.names = 1, stringsAsFactors = F)

# Make PON gender specific for patient and concatenate the data
if(patient.meta$gender == "f") {
  ponXf = pon[pon$Chr == "chrX_F", ]
  ponXf$Chr = "chrX"
  pon = rbind(pon[pon$gender == "neutral", ], ponXf)
  pon$patient = patient[patient$Chr != "chrY", 6]
} else {
  ponXm = pon[pon$Chr == "chrX_M", ]
  ponXm$Chr = "chrX"
  ponYm = pon[pon$Chr == "chrY_M", ]
  ponYm$Chr = "chrY"
  pon = rbind(pon[pon$gender == "neutral", ], ponXm, ponYm)
  pon$patient = patient[, 6]
}
pon$bin = 1:nrow(pon)  # needs recounting after gendering

# Calculate normalized patient counts relative to median sequencing depth and multiply by expected ploidy
pon$patient.norm = pon$patient / median(pon$patient)

# Giemsa-bands for plotting
# The bands table contains the g-bands of the human genome. Cheung VG, Nowak N, Jang W, Kirsch IR, Zhao S, Chen XN, Furey TS, Kim UJ, Kuo WL, Olivier M et al. Integration of cytogenetic landmarks into the draft sequence of the human genome. Nature. 2001 Feb 15;409(6822):953-8. PMID: 11237021
cytobands = read.table("cytobands.tsv", header=TRUE, sep = "\t", row.names = 1, stringsAsFactors = F)
band_types = data.frame(hue=c(0.6, 1, 1, 1, 1, 1, 0.1, 0.8), saturation=c(1, 0, 0, 0, 0, 0, 1, 1), value=c(1, 1, 0.5, 0.8, 0.7, 0.6, 1, 1), row.names=c("acen", "gneg", "gpos100", "gpos25", "gpos50", "gpos75", "gvar", "stalk"))
cytobands$hue = band_types[cytobands$Band.type, "hue"]
cytobands$saturation = band_types[cytobands$Band.type, "saturation"]
cytobands$value = band_types[cytobands$Band.type, "value"]

#begin and end of giemsa bands for karyotyping
#band.borders = read.table("band.borders.tsv", header=TRUE, sep = "\t", row.names = 1, stringsAsFactors = F)

# List of relevant cancer genes
cancer_genes = read.table("cancer_genes.csv", header=TRUE, sep = "\t", stringsAsFactors = F)

# List of all genes
all_genes = read.table("all_genes.tsv", header=TRUE, row.names = 1, stringsAsFactors = F)
all_genes$start = as.numeric(all_genes$start)
all_genes$end = as.numeric(all_genes$end)

# List of chromosomes for plotting
contigs = data.frame(pos=tapply(pon$bin, pon$Chr, mean))
contigs$chr = gsub("chr", "", rownames(contigs))
contigs = contigs[order(as.numeric(contigs$chr)),]
par(mar=c(2, 4, 1, 1))
plot(pon$bin, pon$patient, ylim = c(0, qnorm(0.999, mean(pon$patient), sd(pon$patient))),
     col = "#11111111",
     pch = 16, xaxt="n", xlab="", las = 1, ylab="Read depth", xaxs="i")
abline(v = c(pon$bin[pon$Start == 1], nrow(pon)), col="#55555555", cex = 0.5)
with(contigs[(1:(nrow(contigs)/2))*2, ], axis(side=1, labels=chr, at=pos, line=-.5, cex.axis=0.5, tick=F))
with(contigs[(1:(nrow(contigs)/2))*2-1, ], axis(side=1, labels=chr, at=pos, line=-1, cex.axis=0.5, tick=F))
```


## Sample characteristics and mapping statistics

```{r mapping statistics, echo = FALSE}
stats = data.frame(t(read.table(as.character(patient.meta$mapping_stats), stringsAsFactors = F, row.names=1)))
stats$unique_mapping_pairs = round(sum(patient[ , 6]/2))

stats$raw_read_pairs = format(stats$raw_read_pairs, decimal.mark = ",", big.mark = ".", small.mark = " ", small.interval = 3)
stats$average_read_length = format(stats$average_read_length, decimal.mark = ",", big.mark = ".", small.mark = " ", small.interval = 3)
stats$unique_mapping_pairs = format(stats$unique_mapping_pairs, decimal.mark = ",", big.mark = ".", small.mark = " ", small.interval = 3)

```

| Sample characterisic | Value |
|---------------------:|:------|
| **Sample name**      | **`r as.character(patient.meta$name)`**     |
| **Sample gender**    | **`r as.character(patient.meta$gender)`**   |
| Raw read pairs       | `r as.character(stats$raw_read_pairs)`      |
| Average read length  | `r as.character(stats$average_read_length)` |
| Unique mapping pairs | `r as.character(stats$unique_mapping_pairs)`|


## Distribution of sequencing depth per bin in sample
```{r echo = FALSE, fig.width=7, fig.height=4}
with(pon[!pon$blocked, ], 
     {
      hist(patient, breaks=seq(0, max(patient)+20, by=20),
           xlim=c(0, median(patient)*2),
           freq=F, xlab="Depth per bin", border=T, col="blue", main="")
      pon.hist = hist(norm.clean.median*median(patient), breaks=seq(0, max(norm.clean.median*median(patient)*2,  na.rm=T), by=20), plot=F)
      lines(pon.hist$mids, pon.hist$density, col="red")
      abline(v = c(0.5, 1, 1.5, 2) *median(patient), col='gray', lty=2)
      legend("right", legend="scaled PON", col="red", lty=1, cex=0.8)
      mtext(c("x1", "x2", "x3", "x4"), at=c(0.5, 1, 1.5, 2) *median(patient), side=3)
    })
```

This graph shows the distribution of the bins according to the depth of the sequencing. The red curve represents the distribution of the PON scaled to the same depth. Ideally, the histogram and the red curve match completely. If not, this can indicate larger deletions (the first peak decreases) or larger amplifications (the third peak increases). The indications "x1, x2, x3, x4" indicate the number of copies. 

\newpage

```{r gauss-test, echo = FALSE, fig.width=7, fig.height=3}
sd = mean(unlist(pon$norm.clean.sd), na.rm=T)
pon$pat.z = (pon$patient.norm-1) / sd
pat.dens = density(pon$pat.z, na.rm=T)
plot(seq(from=-20, to=20, by=0.1), dnorm(seq(from=-20, to=20, by=0.1)), type='l', col="gray",
     main="Patient Z-values compared to estimated PON distribution", ylab="Density", xlab="z-value", ylim=c(0, max(pat.dens$y)))
#lines(density((pon$norm.clean.mean-1)/sd, na.rm=T), col="blue")
lines(pat.dens, col="black")
legend("topright", legend=c("theoretical Z-Distribution", "Patient"), col=c("gray", "black"), lty=1, cex=0.7)
  
pon$gauss = apply(pon[, c("patient.norm", "norm.clean.median", "norm.clean.sd")], 1, function(x) ifelse(x[1] == 0, 1, min(pnorm(q=x[1], mean=x[2], sd=x[3], lower.tail=x[1]<x[2])*2, 1)))
pon$gauss.adjust = p.adjust(pon$gauss, method="hochberg")
pon$gauss.adjust[is.na(pon$gauss.adjust)] = 1
par(mar=c(2, 4, 1, 1))
with(pon[!pon$blocked, ],
     plot(bin, patient.norm*2, ylim=c(0, 4),
          col = ifelse(gauss.adjust < 0.05, "#ff000011", "#11111111"),
          pch = 16, xaxt="n", xlab="", las = 1, ylab="ploidy", xaxs="i")
)
# with(pon[!pon$blocked & pon$gauss.adjust < 0.01 , ],
#      arrows(bin, norm.clean.median*2, bin, patient.norm*2, length=0.03, angle=45, code=2, col="#ff000022", lwd=3))
abline(v = c(pon$bin[pon$Start == 1], nrow(pon)), col="#55555555", cex = 0.5)
with(contigs[(1:(nrow(contigs)/2))*2, ], axis(side=1, labels=chr, at=pos, line=-.5, cex.axis=0.5, tick=F))
with(contigs[(1:(nrow(contigs)/2))*2-1, ], axis(side=1, labels=chr, at=pos, line=-1, cex.axis=0.5, tick=F))
```


```{r abberation characters, echo = FALSE}
pon$aberration = factor(ifelse((pon$patient.norm > 1.5 *pon$norm.clean.median & pon$gauss.adjust < 0.05), 'amp', ifelse((pon$patient.norm < 0.75 *pon$norm.clean.median & pon$gauss.adjust < 0.05), 'del', 'normal')), levels = c('normal', 'amp', 'del'))
```

\newpage

## Calculated numerical karyotype:
```{r transform of bins to giemsa bands and create of digital karyotype, echo = FALSE}
cytobands$aberration = factor(apply(cytobands, 1, function(x) {
  bins = pon[pon$Chr == x[1] & pon$Start >= floor(as.integer(x[2])/10^6)*10^6 & pon$End <= ceiling(as.integer(x[3])/10^6)*10^6 & !pon$blocked,]
  if(nrow(bins) < 1) return("normal")
  if(sum(bins$aberration == "del")/nrow(bins) > 0.5) return("del")
  if(sum(bins$aberration == "amp")/nrow(bins) > 0.5) return("amp")
  return("normal")
}))

cytobands$hue[cytobands$aberration == "amp"] = 0.35
cytobands$saturation[cytobands$aberration == "amp"] = 1
cytobands$hue[cytobands$aberration == "del"] = 0
cytobands$saturation[cytobands$aberration == "del"] = 1

cytobands$Band.type[cytobands$aberration == "amp"] = "band add"
cytobands$Color[cytobands$aberration == "amp"] = "darkgreen"

cytobands$Band.type[cytobands$aberration == "del"] = "band del"
cytobands$Color[cytobands$aberration == "del"] = "darkred"

# bands merging
abnorm.bands = cytobands[cytobands$aberration != "normal", ]
karyotype_to_report = abnorm.bands[1, ]
if(nrow(abnorm.bands) < 2) return(karyotype_to_report)

k = 1
# the counter in table karyotype_to_report
for (s in 2:nrow(abnorm.bands)) {
  if (karyotype_to_report$Chr[k] == abnorm.bands$Chr[s] &
      karyotype_to_report$aberration[k] == abnorm.bands$aberration[s] &
      karyotype_to_report$End[k] == abnorm.bands$Start[s]) 
    { karyotype_to_report$last_Name[k] = abnorm.bands$Name[s]
      karyotype_to_report$End[k] = abnorm.bands$End[s]
      next 
    }
  else {
    karyotype_to_report[nrow(karyotype_to_report) +1, ] = abnorm.bands[s, ]
     # die bande auf einem neuen chr oder mit einer anderen richtung hinzufuegen
     k = nrow(karyotype_to_report)
     # das ist jetzt die bande, von der aus betrachtet wird
  }
 }

karyotype_to_report$karyotype = as.character(paste0(karyotype_to_report$aberration, "(",
                                                       sub("chr", "", karyotype_to_report$Chr), ")(",
                                                       karyotype_to_report$Name,
                                                       karyotype_to_report$last_Name, ")"))

kable(karyotype_to_report$karyotype, col.names = NULL, align = "l", "html")
```



\newpage
## Genes affected by CNVs\*:
```{r genes in aberration bins table, echo = FALSE}
  special_genes = all_genes[all_genes$symbol %in% cancer_genes$genes, ]
  special_genes$aberration = 'norm'
  special_genes$found = 0
  
  for(i in 1:nrow(special_genes)) {
    genes.in.bin = pon[(pon$aberration != "normal" & pon$Chr == special_genes$chr[i] & pon$Start <= special_genes$start[i] & special_genes$start[i] <= pon$End) |
                       (pon$aberration != "normal" & pon$Chr == special_genes$chr[i] & pon$Start <= special_genes$end[i] & special_genes$end[i] <= pon$End), ]
    
    if(nrow(genes.in.bin) == 0) {next}
    special_genes$found[i] = 1
    if(all(genes.in.bin$aberration == 'amp')) {special_genes$aberration[i] = 'amp'}
    if(all(genes.in.bin$aberration == 'del')) {special_genes$aberration[i] = 'del'}
  }
  
reported_genes = special_genes[special_genes$found == 1 & (special_genes$aberration == "amp" | special_genes$aberration == "del"), ]
reported_genes$start = format(reported_genes$start, decimal.mark = ",", big.mark = ".", small.mark = " ", small.interval = 3)
reported_genes$end = format(reported_genes$end, decimal.mark = ",", big.mark = ".", small.mark = " ", small.interval = 3)
  
kable(reported_genes[ , 1:6], align = "lllrrl")
```

> Bailey et al. (2018): Comprehensive Characterization of Cancer Driver Genes and Mutations. Cell 173 (2), 371-385.e18. DOI: 10.1016/j.cell.2018.02.060.;

> Papaemmanuil et al. (2016): Genomic Classification and Prognosis in Acute Myeloid Leukemia. The New England journal of medicine 374 (23), S. 2209–2221. DOI:10.1056/NEJMoa1516192.

\newpage

## Chromosome overview plots:

```{r simplePlots, echo = FALSE, fig.width=12, fig.height=3}
units = c('', '', 'K', 'K', 'K', 'M', 'M', 'M')
for(i in unique(pon$Chr)) {
  with(pon[pon$Chr == i,], {
    plot(Start, patient.norm/norm.clean.median *2, col = ifelse(blocked, 0, c("black", "green", "red")[aberration]), ylim=c(0, 4), xaxt='n', xlab='', ylab='Ploidy', pch=20)
    segments(Start, 2+norm.clean.sd*2, Start, 2-norm.clean.sd*2, col="#aaaaaaaa")
      mag = floor(log10(max(End)))
      xmax = floor(max(End)/10^mag)*10^mag
      axis(1, seq(0, xmax/10^mag)*10^mag, paste0(seq(0, xmax/10^mag)*10^(mag%%3), units[mag-mag%%3], ' bp'), line=2)
  })
  mtext(i, side=2, at=-0.75, las=2, cex=1.5)
  abline(h=0:4, lty=3, col="#aaaaaaaa")
  with(cytobands[cytobands$Chr == i, ], {
    rect(Start, -.5, End, -1, xpd=T,
         col=hsv(hue, saturation, value, 0.3)
         )
  })
}
```


\newpage

```{r chromosome plots, echo = FALSE, fig.width=12, fig.height=3}
#plot_bins_regions_ideogram <- function(pon, cytobands, segments, chrn) {

#  x <- pon[pon$Chr == chrn, "Start"]
#  y <- pon[pon$Chr == chrn, "patient.norm"]*2
#  plot_patient <- qplot(x, y, ylim = c(-1, 4)) +
    
#                  geom_line(aes(x = pon[pon$Chr == chrn, "Start"],
#                                y = pon[pon$Chr == chrn, "norm.clean.median"]*2),
#                                color = "gray", linetype = "solid") +
    
#                  geom_errorbar(aes(x = pon[pon$Chr == chrn, "Start"],
#                                    ymin = pon[pon$Chr == chrn, "norm.clean.median"]
#                                           *pon[pon$Chr == chrn, "ploidy"] - 
#                                    pon[pon$Chr == chrn, "norm.clean.sd"]*2, 
#                                    ymax = pon[pon$Chr == chrn, "norm.clean.median"]*2 + 
#                                    pon[pon$Chr == chrn, "norm.clean.sd"]*2),
#                                width=0.1, color = "gray", alpha = 0.9, na.rm = TRUE) +
#                  geom_point(aes(x = x, 
#                                 y = pon[pon$Chr == chrn, "patient.norm"]*2),
#                                 shape = 19, color = "black", size = 1) +
    
                  #das eigentliche ideogram
#                  geom_rect(data = cytobands[cytobands$Chr == chrn, ],
#                            stat = "identity", position = "identity",
#                            aes(NULL, NULL, xmin = Start, xmax = End, fill = Band.type),
#                            ymin = -0.6, ymax = -0.2, color = "gray", size = 0.5, alpha = 0.4) +
    
#                 scale_fill_manual(values=c("acen" = "#2b8cbe", "gneg" = "#f7f7f7",
#                                             "gpos100" = "#252525", "gpos25" = "#cccccc",
#                                             "gpos50" = "#969696", "gpos75" = "#636363",
#                                             "gvar" = "#045a8d", "stalk" = "#be8c2b",
#                                             "amp" = "darkgreen", "del" = "darkred",
#                                             "band add" = "darkgreen", "band del" = "darkred")) +
    
                  # hier legende fuer die banden einfuehren
    
#                  theme_bw() +
#                  labs(title = chrn, y = "Ploidy") +
#                  scale_x_discrete("Giemsa bands")
#  print(plot_patient)
#  return(plot_patient)
#}


#chrn= "chr0"
#  for (chrn in unique(pon$Chr)) {
#    plot_patient = plot_bins_regions_ideogram(pon, cytobands, segments,
#                                              chrn)
#  }
```


The legend contains characteristics to Giemsa stain results. Recognized stain values: gneg, gpos50, gpos75, gpos25, gpos100, acen, gvar, stalk.

> Cheung VG, Nowak N, Jang W, Kirsch IR, Zhao S, Chen XN, Furey TS, Kim UJ, Kuo WL, Olivier M et al. Integration of cytogenetic landmarks into the draft sequence of the human genome. Nature. 2001 Feb 15;409(6822):953-8. PMID: 11237021

