---
title: "coriandR: Panel of Normals"
subtitle: "Report for creation of a new PON"
date: "`r format(Sys.time(), '%d-%m-%Y')`"
output: pdf_document
params:
  pon_name: pon
---

# Generating PON: "`r params$pon_name`"

This report documents the creation of the Panel of Normals (PON) from sequencing data of healthy individuals, which is later used for estimation of calculated karyotype and copy number variations (CNV) in hamatological malignances and solid tumours in research with `coriandR`.

A panel of normals (PON) contains sequencing data from individuals with a normal karyotype who are representative for the analysed population. Multiple samples are used to compensate for or identify technical artefacts and normal biological variability. It is also important that the samples are obtained and processed under the same conditions (same sample preparation methods like DNA extraction and sequencing technology) as the test samples. Using the read depth method, distribution of the reads in samples can be calculated for PON and test samples, which can be used to estimate abberations like CNVs.

```{r, include=FALSE}
library(knitr)

knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
knitr::opts_chunk$set(dev = "png", dpi = 150, dev.args = list(type = "cairo-png"))
```


## Raw Library Counts

Per default raw read counts per library are counted with `featureCounts` for the entire PON at once and stored in the `r params$pon_name`.`fc.tsv` file by the user.

Meta data about the samples (`sample` and `gender`) are specified in the `r params$pon_name`.`meta.tsv` file separated by comma (`,`) by the user.

```{r, data definition, echo = FALSE}
# input meta information with sample mapping files and genders
meta = read.table(paste0(params$pon_name, ".meta.csv"), stringsAsFactors = F, header = TRUE, sep = ",", colClasses = "character")
row.names(meta) = meta$sample

# input raw PON reads counts
# error if PON meta file and featureCounts table have different number or names of samples
pon = read.table(paste0(params$pon_name, ".fc.tsv"), header = TRUE, stringsAsFactors = T)
if(nrow(meta) != ncol(pon)-6) {
  print("PON meta file and featureCounts table have different number of samples")
  knitr::knit_exit()
}

meta$pon.col = as.vector(as.numeric(sapply(meta$sample, function(x) grep(gsub("-", ".", x), names(pon)), simplify=T)))
if(any(is.na(meta$pon.col))) {
  print("PON meta file and featureCounts table have different names of samples")
  knitr::knit_exit()
}

# rename PON colums of samples
names(pon)[meta$pon.col] = meta$sample

# input GC content for human calculated by bedtools
gc = read.table("GRCh38.p13.genome.1M.nucl", stringsAsFactors = T)
# we don`t consider the reads on mitochondrial genome
pon$gc = gc[gc$V1 != 'chrM', 5]
```


### Sequencing depth of individual PON subjects

The following plot shows the sequencing depth in different PON subjects on a log10 scale per megabase (x-axis) for each PON subject with subject ID on the left y-axis and the gender (F/M) from `r params$pon_name`.`meta.tsv` file on the right y-axis. Gender annotation of the samples should match distributions with one single points cluster in females (F) or two distinctive clusters of data points for autosomes and single copy of X and Y chromosomes in males (M).

In optimal case, all samples should be above 100 on x-axis and inside the red dotted lines representing $1/4$ and $4$ times the median sequencing depth overall all PON subjects for control of sequencing depth for each sample.

```{r, sample read counting, warning=F, fig.height=12, fig.width=7, echo = FALSE}
# Display the read distribution per sample in all PON subjects in one plot
par(mar=c(3, 15, 1, 3), cex=0.7)
plot(c(t(pon[ ,meta$pon.col])), rep(1:nrow(meta), nrow(pon)),
     log = "x", xlab="Read depth", ylab="", yaxt = "n",
     pch = 20, col="#11111105")
abline(v=c(0.25, 4)*median(c(t(pon[ ,meta$pon.col])), na.rm=T), col='red', lty=2)
axis(4, 1:nrow(meta), las = 1, meta$gender)
axis(2, 1:nrow(meta), las = 1, meta$sample)
```

  
### Cumulative PON

For later statistical testings the read counts of the PON are summed up for each bin. The reads on autosomal chromosomes are used for the further analysis of both genders, while the reads on the X and Y chromosomes are only used as a statistical reference for the same gender.

Here, read depth is displayed on the y-axis, the genomic coordinates in each chromosome on x-axis. Ideally, the reads located in autosomes are displayed as a straight line of data points with minimal noise above a read depth of at least 1000 reads per megabase (in black colour). Bins with counts outside of 99% of the normal distribution are marked in red colour. These should be masked in the step [Masking bins with extreme GC-content and high variance](#Masking-bins-with-extreme-GC-content-and-high-variance). The values for gonosomes for men and women are represented by (from left to right): X male, X female, Y male, Y female. 

```{r, gender segregation in PON, echo = FALSE, warning=F}
# somatic chromosomes are in "both" PONs, the gonosomes are only in female or male PONs included
pon$ploidy = 2
ponXf = pon[pon$Chr == "chrX", ]
ponXf$Chr = "chrX_F"
ponXf$gender = "F"
ponXf[, meta$sample[meta$gender != "F"]] = NA

ponXm = pon[pon$Chr == "chrX", ]
ponXm$Chr = "chrX_M"
ponXm$gender = "M"
ponXm$ploidy = 1
ponXm[, meta$sample[meta$gender != "M"]] = NA

ponYm = pon[pon$Chr == "chrY", ]
ponYm$Chr = "chrY_M"
ponYm$gender = "M"
ponYm$ploidy = 1
ponYm[, meta$sample[meta$gender != "M"]] = NA

pon$gender = "neutral"

pon = rbind(pon[!pon$Chr %in% c("chrX", "chrY"), ], ponXf, ponXm, ponYm)

pon$gender = factor(pon$gender)

pon$bin = 1:nrow(pon)  # plain number all bins

# chromosome names and middle coordinates of each axis in plot
contigs = data.frame(pos=tapply(pon$bin, pon$Chr, mean))
contigs$chr = gsub("chr", "", rownames(contigs))
contigs = contigs[order(as.numeric(contigs$chr)),]
```


```{r, Cumulative PON, echo = FALSE, fig.width = 12, fig.height = 5}
# One idea is to sum all PON reads of a bin up and compare/test the raw numbers with the patient data. This can be done by binomial testing, or testing the odds ratio.

pon$sum = rowSums(pon[, meta$pon.col], na.rm=T)
pon$mean = rowMeans(pon[, meta$pon.col], na.rm=T)
pon[is.na(pon$mean), "mean"] = 0

plot(pon$bin, pon$mean, ylim = c(0, qnorm(0.999, mean(pon$mean), sd(pon$mean))),
     col = "#11111111", pch = 16, 
     xaxt="n", xlab="", las = 1, ylab="avg. read depth", xaxs="i")
abline(v = c(pon$bin[pon$Start == 1], nrow(pon)), col="#55555555", cex = 0.5)
with(contigs[(1:(nrow(contigs)/2))*2, ], axis(side=1, labels=chr, at=pos, line=1, cex=0.5, tick=F))
with(contigs[(1:(nrow(contigs)/2))*2-1, ], axis(side=1, labels=chr, at=pos, line=0, cex=0.5, tick=F))
```


## Normalising of PON by library size and per megabase

In order to make multiple PONs comparable, read depth is normalised by total library size and then by the median for each megabase. The ploidy of reads located in chromosomes is displayed on the y-axis, the genomic coordinates in each chromosome on x-axis in this plot.

The genome plot should show a distribution of reads around the ploidy of 2. The reads for gonosomes are shifted to meet the ploidy of 2 in males in order to simplify comparability.

```{r, Plot for overview of all chromosomes, echo = FALSE, fig.width = 12, fig.height = 5}
# Normalising of PON reads to a ploidy of 2
pon.norm = pon[, meta$pon.col]
pon.norm = data.frame(t(t(pon.norm)/apply(pon.norm, 2, median, na.rm=T)))
pon$norm.median = apply(pon.norm, 1, median, na.rm=T)
pon$norm.sd = apply(pon.norm, 1, sd, na.rm=T)
pon.nnorm = data.frame(pon.norm/apply(pon.norm, 1, median, na.rm=T))

plot(rep(pon$bin, nrow(meta)), c(as.matrix(pon.nnorm)*pon$ploidy),      
     col ="#11111105", pch = 16, ylim=c(0, 4), yaxs="i",
     xaxt="n", xlab="Chromosomes", las = 1, ylab="Ploidy", xaxs="i")
abline(v=c(pon$bin[pon$Start == 1], nrow(pon)), col="#55555555")
with(contigs[(1:(nrow(contigs)/2))*2, ], axis(side=1, labels=chr, at=pos, line=1, cex=0.5, tick=F))
with(contigs[(1:(nrow(contigs)/2))*2-1, ], axis(side=1, labels=chr, at=pos, line=0, cex=0.5, tick=F))
segments(c(0, 0, min(pon$bin[pon$Chr == "chrX_M"]), min(pon$bin[pon$Chr == "chrX_M"])), c(1.5, 2.5, 1.5, 0.5), c(max(pon$bin[pon$Chr == "chrX_F"]), max(pon$bin[pon$Chr == "chrX_F"]), max(pon$bin), max(pon$bin)), col="red", lty=2)
axis(4, 0:4, 0:4, las=2)
```

\newpage

## Masking bins with extreme gc content and high variance

To improve the quality of the PON and to minimize technical artifacts, we identify the bins with a gc content that deviates from the expected value for the human genome. We determined the gc content of the human reference genome using Bedtools v2.27.1 (Quinlan 2014).

**Plot of the gc content (x-axis) against variance (left plot) and ploidy (right plot) in normalised bin counts in PON:**

Bins with a gc content less than the 0.275 percentile genome-wide assumed normal distribution are considered to be extreme, marked in red colour in the plot and masked from the PON, that means that they are not considered for further statistical testings.

- If gc content has a slight influence on variance of reads, a cigar-shaped cluster of data points appears at the ploidy of 2. 
- If gc content has a high influence on variance of reads, a diagonal or a banana-shaped cluster of data points can be observed. It may indicate a technical artifact in library preparation. 

**References:**

Quinlan A. R. (2014). BEDTools: The Swiss-Army Tool for Genome Feature Analysis. Current protocols in bioinformatics, 47, 11.12.1–11.12.34. doi: 10.1002/0471250953.bi1112s47.

```{r, plot GC-content, echo = FALSE, fig.width = 7, fig.height = 3}
pon$nnorm.var = c(apply(pon.nnorm, 1, var, na.rm=T))
pon$gc_blocked = pon$gc < qnorm(0.275, mean(pon$gc, na.rm=T), sd(pon$gc, na.rm=T))
par(mfrow=c(1,2))

plot(pon$gc, pon$nnorm.var, pch=20, ylim=c(0.0001, 0.2), xlim=c(0.25, 0.6),
     col=ifelse(pon$gc_blocked, "tomato3", "#11111111"),
     xlab="gc content", ylab="variance", las = 1, cex = 0.5)
points(rep(0.25, sum(pon$gc < 0.25)), pon$nnorm.var[pon$gc < 0.25], pch="<", col="tomato3")
points(pon$gc[!is.na(pon$nnorm.var) & pon$nnorm.var > 0.2], rep(0.2, sum(!is.na(pon$nnorm.var) & pon$nnorm.var > 0.2)), pch="^", col=ifelse(pon$gc_blocked, "tomato3", "#11111122")[!is.na(pon$nnorm.var) & pon$nnorm.var > 0.2])
gc.var = predict(loess(nnorm.var ~ gc, pon[pon$gc > 0.35 & pon$gc < 0.6,], span=0.5), 350:600/1000)
lines(350:600/1000, gc.var, col="blue")
# Bins with a GC-content < the 0.275 percentile genome-wide assumed normal distribution or with variance > 0.2 were considered to be extreme and masked from the PON

plot(rep(pon$gc, ncol(pon.norm)), c(as.matrix(pon.norm)*pon$ploidy), pch=20, xlim=c(0.25, 0.6), ylim=c(0, 4),
     col = ifelse(pon$gc_blocked, "tomato3", "#11111111"),
     xlab="gc content", ylab="ploidy", las = 1, cex = 0.5)
abline(h=c(1.5, 2.5), col="red", lty=2)
```

# Masking of bins with a high variance, deviating gc content or both

The ploidy of reads located in chromosomes is displayed on the y-axis, the genomic coordinates in each chromosome on x-axis. This plot displays the bins (a window with size of one megabasewith counted reads in it) that are excluded from the statistical testings in different colours: very low GC-content (red), high variance (blue), or both (purple). Those bins are often located in centromeres and telomeres, both regions with high percentage of repetitive DNA regions.

```{r, Masking high variance bins, echo = FALSE, fig.width = 12, fig.height = 5}
# One idea would be to clear single bins of single PON subjects where depth is below or above 99% of other subjects for that bin.
# For small PONs this is to coarse and probably not warrented overall...

norm.sd = mean(apply(pon.norm, 1, sd, na.rm=T))
pon.norm.clean = data.frame(t(apply(pon.norm, 1, function(x) {x[x < qnorm(0.01, mean(x), norm.sd) | x > qnorm(0.99, mean(x), norm.sd)] = NA; x})))
pon$norm.clean.mean = rowMeans(pon.norm.clean, na.rm=T)
pon$norm.clean.median = c(apply(pon.norm.clean, 1, median, na.rm=T))
pon$norm.clean.var = c(apply(pon.norm.clean, 1, var, na.rm=T))
pon$norm.clean.sd = c(apply(pon.norm.clean, 1, sd, na.rm=T))
pon$norm.clean.iqr = c(apply(pon.norm.clean, 1, IQR, na.rm=T))
pon$var_blocked = pon$sum < 1 | (!is.na(pon$norm.clean.var) & pon$norm.clean.var > qnorm(0.99, mean(pon$norm.clean.var[!pon$gc_blocked], na.rm=T), sd(pon$norm.clean.var[!pon$gc_blocked], na.rm=T)))
pon$blocked = pon$var_blocked | pon$gc_blocked

plot(rep(pon$bin, nrow(meta)), c(as.matrix(pon.norm.clean)*2),
     col=ifelse(pon$gc_blocked, ifelse(pon$var_blocked, "#aa33aa55", "#aa331155"), ifelse(pon$var_blocked, "#1133aa55", "#11111105")),
     ylim=c(0, 4), yaxs="i", pch=16,
     xaxt="n", xlab="Chromosomes", las = 1, ylab="Ploidy", xaxs="i")
abline(v = c(pon$bin[pon$Start == 1], nrow(pon)), col="#55555555", cex = 0.5)
with(contigs[(1:(nrow(contigs)/2))*2, ], axis(side=1, labels=chr, at=pos, line=1, cex=0.5, tick=F))
with(contigs[(1:(nrow(contigs)/2))*2-1, ], axis(side=1, labels=chr, at=pos, line=0, cex=0.5, tick=F))
segments(c(0, 0, min(pon$bin[pon$Chr == "chrX_M"]), min(pon$bin[pon$Chr == "chrX_M"])), c(1.5, 2.5, 1.5, 0.5), c(max(pon$bin[pon$Chr == "chrX_F"]), max(pon$bin[pon$Chr == "chrX_F"]), max(pon$bin), max(pon$bin)), col="red", lty=2)
axis(4, 0:4, 0:4, las=2)
```


```{r, Overview masked bins, echo = FALSE, fig.width = 12, fig.height = 5, warning=F}
plot(pon$bin, pon$norm.clean.median*2,
     col=ifelse(pon$gc_blocked, ifelse(pon$var_blocked, "#aa33aa55", "#aa331155"), ifelse(pon$var_blocked, "#1133aa55", "#11111105")),
     ylim=c(0, 4), yaxs="i", pch=16,
     xaxt="n", xlab="Chromosomes", las = 1, ylab="Ploidy", xaxs="i")
arrows(pon$bin, (pon$norm.clean.median+pon$norm.clean.sd)*2, pon$bin, (pon$norm.clean.median-pon$norm.clean.sd)*2, length=0.01, angle=90, code=3, col=ifelse(pon$gc_blocked, ifelse(pon$var_blocked, "#aa33aa99", "#aa331199"), ifelse(pon$var_blocked, "#1133aa99", "#11111105")))
abline(v = c(pon$bin[pon$Start == 1], nrow(pon)), col="#55555555", cex = 0.5)
with(contigs[(1:(nrow(contigs)/2))*2, ], axis(side=1, labels=chr, at=pos, line=1, cex=0.5, tick=F))
with(contigs[(1:(nrow(contigs)/2))*2-1, ], axis(side=1, labels=chr, at=pos, line=0, cex=0.5, tick=F))
segments(c(0, 0, min(pon$bin[pon$Chr == "chrX_M"]), min(pon$bin[pon$Chr == "chrX_M"])), c(1.5, 2.5, 1.5, 0.5), c(max(pon$bin[pon$Chr == "chrX_F"]), max(pon$bin[pon$Chr == "chrX_F"]), max(pon$bin), max(pon$bin)), col="red", lty=2)
axis(4, 0:4, 0:4, las=2)
```


\newpage

# gc content in PON after masking of bins with deviating gc content

This plot displays the gc content (x-axis) against variance (left plot) and ploidy (right plot) in normalised bin counts in PON after masking of bins with deviating gc content. gc content should not have any influence on the ploidy of the bins in chromosomes.

```{r, gc content after correction, echo = FALSE, fig.width = 7, fig.height = 3}
par(mfrow=c(1,2))
with(pon[!pon$blocked, ], {
plot(gc, norm.clean.var, col="#11111105", pch=20,
     ylim=c(0.00001, 0.2), xlim=c(0.25, 0.6),
     xlab="gc content", ylab="variance", las = 1, cex = 0.5)
gc.var.clean = predict(loess(norm.clean.var ~ gc, pon[pon$gc > 0.33 & pon$gc < 0.6,], span=0.5), 330:550/1000)
lines(330:550/1000, gc.var.clean, col="blue")
plot(rep(gc, ncol(pon.norm)), c(as.matrix(pon.norm.clean[!pon$blocked, ])*pon$ploidy), pch=20, xlim=c(0.25, 0.6), ylim=c(0, 4), col="#11111105",
     xlab="gc content", ylab="Ploidy", las = 1, cex = 0.5)
abline(h=c(1.5, 2.5), col="red", lty=2)
})
```


# Overview over the final PON after masking of bins with deviating gc content, high variance or both

The ploidy of reads located in chromosomes is displayed on the y-axis, the genomic coordinates in each chromosome on x-axis. The bins in PON should be located closely around the ploidy of 2.

```{r, corrected PON overview, echo = FALSE, fig.width = 12, fig.height = 5}
plot(rep(pon$bin[!pon$blocked], nrow(meta)), c(as.matrix(pon.norm.clean[!pon$blocked, ])*2),
     col ="#11111105", pch = 16, ylim=c(0, 4), yaxs="i",
     xaxt="n", xlab="Chromosomes", las = 1, ylab="Ploidy", xaxs="i")
abline(v = c(pon$bin[pon$Start == 1], nrow(pon)), col="#55555555", cex = 0.5)
with(contigs[(1:(nrow(contigs)/2))*2, ], axis(side=1, labels=chr, at=pos, line=1, cex=0.5, tick=F))
with(contigs[(1:(nrow(contigs)/2))*2-1, ], axis(side=1, labels=chr, at=pos, line=0, cex=0.5, tick=F))
segments(c(0, 0, min(pon$bin[pon$Chr == "chrX_M"]), min(pon$bin[pon$Chr == "chrX_M"])), c(1.5, 2.5, 1.5, 0.5), c(max(pon$bin[pon$Chr == "chrX_F"]), max(pon$bin[pon$Chr == "chrX_F"]), max(pon$bin), max(pon$bin)), col="red", lty=2)
axis(4, 0:4, 0:4, las=2)
```


```{r, save the PON table, echo = FALSE}
write.table(pon, file = paste0(params$pon_name, ".pon.tsv"), sep = "\t", row.names = TRUE, col.names = NA)
```
